---
title: "Getting started with qmdme"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with qmdme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

`qmdme` produces and updates `.qmd` companion files for the source code in
a research project. Each source file (`R/fit.R`, `sql/query.sql`, ...) gets
a corresponding `qmd/<mirror>/<name>.qmd` with a user-editable prose
region above an `eval=FALSE` code chunk. The chunk shows the source code
verbatim; nothing executes during render.

## Setup

```{r}
library(qmdme)
qmdme::init()
```

`init()` creates `qmd/` and a stub `qmd/index.qmd`. This default
(`scope = "embed"`) assumes the companions ride along inside a Quarto site you
already maintain, so `init()` looks at your project root, works out what kind of
site you have, and tells you the next step — typically "run `qmdme::wire()` to
connect `qmd/` to it" (see
[Connecting to an existing site](#connecting-to-an-existing-site) below).

If you'd rather have `qmdme` build the site for you, pass `scope = "website"`:

```{r}
qmdme::init(scope = "website")
```

This scaffolds a self-contained, navigable Quarto website rooted at `qmd/`:

- `qmd/_quarto.yml` — `project: type: website` with a navbar (Home / About) and
  an auto sidebar that lists every companion `sync()` writes, so navigation
  stays current with no manual edits. It renders to `qmd/_site` and sets
  `execute: enabled: false`, so a render never starts a language kernel (the
  companions are display-only `eval=FALSE` chunks).
- `qmd/index.qmd` — a landing page.
- `qmd/about.qmd` — a starter page describing how the companions are generated.

Every seed file is skip-if-exists, so re-running `init()` never clobbers your
edits, and `scope = "website"` never touches a `_quarto.yml` at the project
root.

## Generating companions

```{r}
qmdme::sync()
```

`sync()` walks the project for `.R`, `.sql`, and `.py` files (skipping
`.git`, `renv`, `.venv`, the `qmd/` output, and other standard ignored
paths), then writes one companion file per source.

Narrow scope by passing `paths`:

```{r}
qmdme::sync(paths = "R/")
```

## Connecting to an existing site

If you already maintain a Quarto site, you don't want to go spelunking through
YAML to find where `qmd/` plugs in. `wire()` does the edit for you: it looks at
the project root and, when an edit is needed, points your root `_quarto.yml` at
the `qmd/` folder.

```{r}
qmdme::wire()
```

`wire()` edits **only when it has to**, which keeps it honest about how Quarto
works:

* If the site has an explicit sidebar `contents:` list that omits `qmd`,
  `wire()` adds `qmd` to it. Same for an explicit `project: render:` allowlist.
* If the site already surfaces `qmd` automatically — an `auto` sidebar or no
  `render:` restriction, which is the common minimal config — there's nothing to
  do, and `wire()` says so without touching the file.
* If `qmd/` is already a standalone site (you used `scope = "website"`), or the
  site is pkgdown / R Markdown, or there's no site at all, `wire()` doesn't edit
  anything; it explains and points you at `init(scope = "website")`.

Because the edit is a `yaml` read-modify-write, it does not preserve comments or
exact formatting in your `_quarto.yml`. `wire()` therefore backs the file up to
`_quarto.yml.bak` before editing (pass `backup = FALSE` to skip), and if your
config is heavily commented you may prefer to add `qmd` by hand. `wire()`
returns its result invisibly (`type`, `file`, `changed`, `backup`) so you can
branch on it in a script.

`sync()` will also remind you, with a warning, when the companions aren't
connected to a site yet. If you don't want a site, silence it with
`options(qmdme.warn_no_site = FALSE)` or `sync(warn_no_site = FALSE)`.

## Editing companions

Each generated `.qmd` looks like this:

````
---
title: "fit.R"
---

## Notes

<!-- everything above the marker is yours; sync() never touches it -->

<!-- qmdme:code-below — do not edit below this line; regenerated by
     sync() on 2026-05-01 14:32 -->

```{r, eval=FALSE}
fit_model <- function(data) { ... }
```
````

Edit anything above the sentinel — add prose, headings, links — and
re-run `sync()`. Your edits stay; the chunk below the sentinel is
refreshed from source.

If the sentinel is missing (for example, you accidentally deleted it),
`sync()` re-appends a fresh code section at the bottom and warns so
you notice the file was repaired. Companion files always end up with
one current sentinel + chunk.

## Rendering the website

If you scaffolded with `init(scope = "website")`, `qmd/` is a complete Quarto
site. After syncing, build it from the shell:

```sh
quarto render qmd
```

The rendered HTML lands in `qmd/_site`, with the landing page, the About page,
and one page per companion, all reachable from the navbar and sidebar. Re-run
`qmdme::sync()` whenever sources change, then `quarto render qmd` again — new
companions appear in the sidebar automatically.
