---
title: "Getting started with qmdme"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with qmdme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

`qmdme` produces and updates `.qmd` companion files for the source code in
a research project. Each source file (`R/fit.R`, `sql/query.sql`, ...) gets
a corresponding `qmd/<mirror>/<name>.qmd` with a user-editable prose
region above an `eval=FALSE` code chunk. The chunk shows the source code
verbatim; nothing executes during render.

## Setup

```{r}
library(qmdme)
qmdme::init()
```

`init()` creates `qmd/` and a stub `qmd/index.qmd`. This default
(`scope = "embed"`) assumes the companions ride along inside a Quarto site you
already maintain. To find out exactly how to connect `qmd/` to that site, run
`wire()` (see [Connecting to an existing site](#connecting-to-an-existing-site)
below).

If you'd rather have `qmdme` build the site for you, pass `scope = "website"`:

```{r}
qmdme::init(scope = "website")
```

This scaffolds a self-contained, navigable Quarto website rooted at `qmd/`:

- `qmd/_quarto.yml` — `project: type: website` with a navbar (Home / About) and
  an auto sidebar that lists every companion `sync()` writes, so navigation
  stays current with no manual edits. It renders to `qmd/_site` and sets
  `execute: enabled: false`, so a render never starts a language kernel (the
  companions are display-only `eval=FALSE` chunks).
- `qmd/index.qmd` — a landing page.
- `qmd/about.qmd` — a starter page describing how the companions are generated.

Every seed file is skip-if-exists, so re-running `init()` never clobbers your
edits, and `scope = "website"` never touches a `_quarto.yml` at the project
root.

## Generating companions

```{r}
qmdme::sync()
```

`sync()` walks the project for `.R`, `.sql`, and `.py` files (skipping
`.git`, `renv`, `.venv`, the `qmd/` output, and other standard ignored
paths), then writes one companion file per source.

Narrow scope by passing `paths`:

```{r}
qmdme::sync(paths = "R/")
```

## Connecting to an existing site

If you already maintain a Quarto site, you don't want to go spelunking through
YAML to find where `qmd/` plugs in. `wire()` does that for you: it looks at the
project root, works out what kind of site you have, and tells you exactly which
file to edit and which key to add `qmd` under.

```{r}
qmdme::wire()
```

* If it finds a root `_quarto.yml`, it prints the precise snippet to add (and
  notes when `qmd` is already referenced, so re-running is safe).
* If `qmd/` is already a standalone site (you used `scope = "website"`), it
  reminds you to `quarto render qmd` -- there's nothing to wire.
* If it finds a pkgdown or R Markdown site (which don't build from loose `.qmd`
  files), or no site at all, it points you at `init(scope = "website")`.

`wire()` is read-only -- it never edits or creates files, so it can't corrupt a
config you maintain by hand. It also returns its findings invisibly
(`type`, `file`, `wired`) if you want to branch on them in a script.

## Editing companions

Each generated `.qmd` looks like this:

````
---
title: "fit.R"
---

## Notes

<!-- everything above the marker is yours; sync() never touches it -->

<!-- qmdme:code-below — do not edit below this line; regenerated by
     sync() on 2026-05-01 14:32 -->

```{r, eval=FALSE}
fit_model <- function(data) { ... }
```
````

Edit anything above the sentinel — add prose, headings, links — and
re-run `sync()`. Your edits stay; the chunk below the sentinel is
refreshed from source.

If the sentinel is missing (for example, you accidentally deleted it),
`sync()` re-appends a fresh code section at the bottom and warns so
you notice the file was repaired. Companion files always end up with
one current sentinel + chunk.

## Rendering the website

If you scaffolded with `init(scope = "website")`, `qmd/` is a complete Quarto
site. After syncing, build it from the shell:

```sh
quarto render qmd
```

The rendered HTML lands in `qmd/_site`, with the landing page, the About page,
and one page per companion, all reachable from the navbar and sidebar. Re-run
`qmdme::sync()` whenever sources change, then `quarto render qmd` again — new
companions appear in the sidebar automatically.
